% Prof. Sumi & committee
I owe my deepest gratitude to my advisor Professor Yasuyuki Sumi, who gave me the opportunity to join his group and from whom I learned a lot on this Ph.D. journey. On our first meeting, you gave me the freedom to decide my line of research. I may not have always gone to you for advice because I always try to figure out and do things on my own first, but when I do, you never fail to give deep level of insight that inspired significant improvements on my work. Working with you, I have always appreciated how you elevated my simple ideas and challenged my over-ambitious ones. I also value how you always remind me to communicate my ideas simpler and better. Thank you for everything.

% Labmates - Kaisei, Okuno, Fujikura, Takasaki, Suzuki, Shiohara, Watanabe, Murakami, ohashi, maeno, tada, hiroki yamamoto, azuma, toyama, takuya takahashi, kitamura, 
My Ph.D. journey would not have been the same without the support of the masters and bachelor students in our lab who helped in improving my study designs and in helping me navigate Japanese life and bureaucracies. They were my buddies and some, I would consider my close friends. I would like to thank Kaisei Tsujimoto for teaching me conversational Japanese and being a good friend, Ryo Fujikura, Takuya Takahashi, Yuki Watanabe, Hiroki Yamamoto, and Akane Okuno for their tremendous help when I was adjusting as a new member of the lab, and Masahiro Takasaki, Kiyomasa Murakami, Koki Shiohara, Ryo Suzuki, Naruse Maeno, Ayaka Tada, Kai Toyama, Kuraaku Azuma, and Kazuma Ohashi for helping me in my research during the last years.

% COMET colleagues and members
Much of my work was done remotely in the Philippines and the success of my studies are partly because of the great help from my colleagues and student members of the Center for Complexity and Emerging Technologies at De La Salle University. In particular, I would like to thank Luigi Acorda and Jasper Pillejera for helping me conduct my formative study with drivers, and Unisse Chua for helping me recruit participants in my last few studies. 

% FUN Staff
I also want to thank all the Japanese language teachers at the youth center, especially Katsuko Takahashi, for voluntarily teaching me basic Japanese. They have equipped me with everything I need to survive everyday living in Japan.

% Filipino community in Hakodate
Going into a Ph.D. program puts you in a very stressful, frustrating, and sometimes isolating environment. I managed to live a balanced social life in Japan because of the great community provided other Filipinos that I met through church and other scholars. In particular, I would like to thank Joemark Narsico for connecting me with the Filipino community, Maricel Oya, Maria Hikosaka and Divine De Padua Asai for their overflowing generosity, care and hospitality, and Delia, Marigold, Maria, Haru, Aireen, Reynel, Jon, Jay, Yuki, Lei-ya, Michie, and Jeralyn for the countless parties and outings. I would also like to thank the Filipino scholars in Hokkaido University for always hosting me in Sapporo whenever I attend their monthly meetups. In particular, I would like to thank Janine Tolod, Danjo De Chavez, Kunihito Mihara, Ianne Ubalde, Jin Algodon, Jacob Dy, Lawrence Belotindos, Ren, Ellison Castro, and Ronald Reyes for their generous hospitality.

% VGang
I also want to thank my close group of friends, Jamie Dela Cruz for always hosting me in Tokyo whenever I go to conferences and visa interviews, Angel Directo, Davidson Oliveros, JM Sta. Juana, Eldin Lao, Janelle Kintanar, Dynna De Asis, Stephanie Seranillo, Alyssa Bueno, Macy Bautista, Nadine Camunary, Lorraine Cueva, Neezil Escol, Ellen Deferia, Andrew Lachica, Kevin Crisologo, Erwin Casia, Seigfrid Mendoza, and Therese Co for supporting me in various stages of my Ph.D. and have kept me company during the pandemic. 

% Parents and family
Finally, I want to thank my parents Maria Cristina and Roberto Samson, and siblings Fionna and Robert for their unwavering support and for never stopping me to pursue my academic career. Thank you for always making sure that my infrequent visits to the Philippines are as worry free as possible. Thank you for always filling my bags with Filipino food items and souvenirs when I leave. 